|

### 2024-12-16 manuscript_tf/manuscript.Rmd
AE

---
title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
# author:
#   - name: Damien Dupré
#     email: damien.dupre@dcu.ie
#     affiliation: Business School, Dublin City University
#     correspondingauthor: true
#   - name: Edgar Morgenroth
#     email: edgar.morgenroth@dcu.ie
#     affiliation: Business School, Dublin City University
# address:
#   - code: Business School, Dublin City University
#     organization: Business School, Dublin City University
#     city: Dublin
#     country: Ireland
abstract: |
  *Objectives:* This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.

  *Study Design:* A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.

  *Methods:* A Generalized Additive Model is performed to compare trends from twelve European countries across five age groups. Then, Transfer Entropy calculations are used to evaluate the impact of younger cohorts on older cohorts.

  *Results:* The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.

  *Conclusions:* These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
keywords: 
  - COVID-19
  - School Closures
  - Intergenerational Transmission.
journal: "Public Health"
date: "`r Sys.Date()`"
classoption: preprint, 3p
linenumbers: false
numbersections: false
bibliography: bibliography.bib
# biblio-style: elsarticle-num
csl: https://www.zotero.org/styles/elsevier-with-titles
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: false
    citation_package: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---

(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs. closure due to government decisions. Source: @unesco2022https.
(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country.
(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group.
(ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country. Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$). The significance of each Transfer Entropy coefficient is provided in Appendix 2.

@@

A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines. Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial]. It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19. They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.

Recently, a body of literature has shown various negative consequences on child development and health due to school closures. School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading]. Some governments are now questioning their decisions to close schools [@de2021determines]. For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach]. Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19. However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive. @alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence. In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity. The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.

@@

Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project [@riffe2021data]. The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available. In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country.

To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022). For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break. Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered. Indeed, both are considered as closure at a country-wide level. Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days. Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.

@@

The effect of school closure on the trend of daily COVID-19 cases is analyzed using a Generalized Additive Model (GAM). GAM is a flexible modeling approach that estimates non-linear relationships between variables. Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.

@@

\begin{align*}
  \tag{1}
  n\prime_t &\sim Poisson(\mu_t) \\
  \log(\mu_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t + \\
              & f_1(closure_t) + f_2(closure_t,country_t) + f_3(closure_t,age\,group_t)
\end{align*}

\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases. The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups. The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group. The restricted maximum likelihood (REML) was used to avoid overfitting while estimating smoothing parameters. In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model. By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.

@@

Transfer entropy ($T$) can be used to infer the temporal relationship between two time series $X$ and $Y$. This measure indicates whether changes in COVID-19 cases within a certain age group $X$ can be used to reduce the uncertainty on the future COVID-19 cases within another age group $Y$. If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring]. As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger]. Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].

@@

\begin{align}
  \tag{2}
  T_{ag\,x \rightarrow\,ag\,y}(k,l) = \sum_{ag\,x,ag\,y} p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) \log \left(\frac{p\left(ag\,x_{t+1}| ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1}|ag\,x_t^{(k)}\right)}\right)
\end{align}

@@

Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)). For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial two weeks following school closure, followed by flattening off (`r gam_results["0to4"]`). In contrast, for the age group ranging from 5 to 14, a substantial increase in cases after 14 days of school closure is found (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`). For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`). Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.

@@

Knowledge about the transmission of the virus significantly improved over time as more studies have been published. While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate. This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups. While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups. Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization. In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure. Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days. Thus, the results do not support the hypothesis that school closures were effective. These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.

The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y). Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence. This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional. These findings highlight the importance of considering intergenerational interactions in designing effective control measures.

Furthermore, the quadrant analysis reveals similar patterns across multiple countries. Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts. On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.

In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures. While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case.

@@

<!-- <div id="refs"></div> -->

<!-- \newpage -->

<!-- # Appendix 1: Dickey-Fuller test for stationay timeseries {-}  -->

<!-- Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series). -->

<!-- The Augmented Dickey-Fuller test incorporates three types of linear regression models. The first type is a linear model with no drift and linear trend with respect to time. The second type is a linear model with drift but no linear trend. The third type is a linear model with both drift and linear trend. -->

<!-- The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.  -->

<!-- ```{r} -->
<!-- adf_res_type1 <- df_global_age |> -->
<!--   group_by(age_grp, country) |> -->
<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
<!--                  bind_rows() |>  -->
<!--                  pluck("type1") |>  -->
<!--                  as_tibble() -->
<!--   ) |>  -->
<!--   rename(ADF_type1 = ADF, p.value_type1 = p.value) -->

<!-- adf_res_type2 <- df_global_age |> -->
<!--   group_by(age_grp, country) |> -->
<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
<!--                  bind_rows() |>  -->
<!--                  pluck("type2") |>  -->
<!--                  as_tibble() -->
<!--   ) |>  -->
<!--   rename(ADF_type2 = ADF, p.value_type2 = p.value) -->

<!-- adf_res_type3 <- df_global_age |> -->
<!--   group_by(age_grp, country) |> -->
<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
<!--                  bind_rows() |>  -->
<!--                  pluck("type3") |>  -->
<!--                  as_tibble() -->
<!--   ) |>  -->
<!--   rename(ADF_type3 = ADF, p.value_type3 = p.value) -->

<!-- adf_res <- adf_res_type1 |>  -->
<!--   full_join(adf_res_type2) |>  -->
<!--   full_join(adf_res_type3) -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- adf_res |> -->
<!--   filter(age_grp %in% c("aged0to4", "aged5to9")) |>  -->
<!--   kbl( -->
<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
<!--   collapse_rows(columns = 1:2, valign = "middle") -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- adf_res |> -->
<!--   filter(age_grp %in% c("aged10to14", "aged15to19")) |>  -->
<!--   kbl( -->
<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
<!--   collapse_rows(columns = 1:2, valign = "middle") -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- adf_res |> -->
<!--   filter(age_grp %in% c("aged20to24", "aged25to34")) |>  -->
<!--   kbl( -->
<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
<!--   collapse_rows(columns = 1:2, valign = "middle") -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- adf_res |> -->
<!--   filter(age_grp %in% c("aged35to44", "aged45to54")) |>  -->
<!--   kbl( -->
<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
<!--   collapse_rows(columns = 1:2, valign = "middle") -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- adf_res |> -->
<!--   filter(age_grp %in% c("aged55to64", "aged65+")) |>  -->
<!--   kbl( -->
<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
<!--   collapse_rows(columns = 1:2, valign = "middle") -->
<!-- ``` -->

<!-- \newpage -->

<!-- # Appendix 2: Significance of Transfer Entropy coefficients -->

<!-- Results for each permutation of age groups from each country. Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence. -->

<!-- ```{r} -->
<!-- # variables -------------------------------------------------------------------- -->
<!-- age_order <- c( -->
<!--   "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24", -->
<!--   "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+" -->
<!-- ) -->

<!-- # data ------------------------------------------------------------------------- -->
<!-- te_results <- read_csv(here("data/te_results.csv")) |>  -->
<!--     mutate( -->
<!--       X = factor(X, levels = age_order), -->
<!--       Y = factor(Y, levels = age_order) -->
<!--     ) |>  -->
<!--   select(-direction) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- te_results |> -->
<!--   filter(X == "aged0to4") |>  -->
<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
<!--   arrange(Y) |>  -->
<!--   kbl( -->
<!--     caption = "Significance of Transfer Entropy coefficients", -->
<!--     digits = 2 -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
<!--   kable_styling(font_size = 7) -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- te_results |> -->
<!--   filter(X == "aged0to4") |>  -->
<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
<!--   arrange(Y) |>  -->
<!--   kbl( -->
<!--     caption = "Significance of Transfer Entropy coefficients", -->
<!--     digits = 2 -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
<!--   kable_styling(font_size = 7) -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- te_results |> -->
<!--   filter(X == "aged5to9") |>  -->
<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
<!--   arrange(Y) |>  -->
<!--   kbl( -->
<!--     caption = "Significance of Transfer Entropy coefficients", -->
<!--     digits = 2 -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
<!--   kable_styling(font_size = 7) -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- te_results |> -->
<!--   filter(X == "aged5to9") |>  -->
<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
<!--   arrange(Y) |>  -->
<!--   kbl( -->
<!--     caption = "Significance of Transfer Entropy coefficients", -->
<!--     digits = 2 -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
<!--   kable_styling(font_size = 7) -->
<!-- ``` -->

<!-- \newpage -->
<!-- ```{r} -->
<!-- te_results |> -->
<!--   filter(X == "aged10to14") |>  -->
<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
<!--   arrange(Y) |>  -->
<!--   kbl( -->
<!--     caption = "Significance of Transfer Entropy coefficients", -->
<!--     digits = 2 -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
<!--   kable_styling(font_size = 7) -->
<!-- ``` -->

<!-- \newpage -->

<!-- ```{r} -->
<!-- te_results |> -->
<!--   filter(X == "aged10to14") |>  -->
<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
<!--   arrange(Y) |>  -->
<!--   kbl( -->
<!--     caption = "Significance of Transfer Entropy coefficients", -->
<!--     digits = 2 -->
<!--   ) |>  -->
<!--   kable_classic(full_width = F) |>  -->
<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
<!--   kable_styling(font_size = 7) -->
<!-- ``` -->

<!-- \newpage -->

>>>>

---
title: |
  Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
author:
  - name: Damien Dupré
    affil: 1
    email: damien.dupre@dcu.ie
  - name: Edgar Morgenroth
    affil: 1
    email: edgar.morgenroth@dcu.ie
affiliation:
  - num: 1
    address: |
      Business School, Dublin City University

# author:
#   - name: Author 1
#     affil: 1
#     email: Email 1
#   - name: Author 2
#     affil: 2
#     email: Email 2
# affiliation:
#   - num: 1
#     address: |
#       Affiliation 1
#   - num: 2
#     address: |
#       Affiliation 2

bibliography: bibliography.bib
abstract: |
  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups. The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM). The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa. These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
keywords: |
  COVID-19; School Closures; Intergenerational Transmission.
output:
  bookdown::pdf_book:
    base_format: rticles::tf_article
    keep_tex: false
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
  - \usepackage{hyperref}
  - \usepackage[utf8]{inputenc}
  - \def\tightlist{}
---

(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions. Source: @unesco2022https.
(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries.
(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups.
(ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country. Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$). The significance of each Transfer Entropy coefficient is provided in Appendix 2.

@@

The COVID-19 pandemic has had a profound impact on global health, with an estimate of 14.83 million excess deaths globally [@msemburi2023estimates]. Beyond direct mortality, the pandemic has caused significant collateral damage, including losses of lives and livelihoods, necessitating a comprehensive approach to measuring its broader impacts. A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines. Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial]. It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19. They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.

Recently, a body of literature has shown various negative consequences on child development and health due to school closures. School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading]. Some governments were questioning their decisions to close schools [@de2021determines]. For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach]. Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19. However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive. In a study of a panel of European countries, @alfano2022effects found that school closures were associated with a reduced COVID-19 incidence. In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity. The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.

@@

Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project [@riffe2021data]. The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available. In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country. Countries communicating data with groups of 5 years were chosen to match as much as possible the different school stages.

To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022). For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break. Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered. Indeed, both are considered as closure at a country-wide level. Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.

@@

The effect of school closure on the trend of daily COVID-19 cases is analysed using a Generalized Additive Model (GAM). GAM is a flexible modelling approach that estimates non-linear relationships between variables. Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.

@@

\begin{align*}
  \tag{1}
  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
  &\quad + f_3(closure_t, age\,group_t) + \epsilon_t
\end{align*}

\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases. The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups. The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group. The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters. In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as $\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)$ where $\phi$ is the autoregressive parameter. By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.

Chi-square statistics were employed to assess if the degree of smoothness is significantly different from zero.

@@

Transfer entropy ($T$) is a measure of the directional information flow between two time series $X$ and $Y$, capturing how the past values of one variable can predict changes in another. Unlike correlation, transfer entropy accounts for the temporal order of events and non-linear relationships, making it particularly suited for studying dynamic systems such as epidemic data. In this analysis, transfer entropy quantifies how case counts in one age group $X$ influence subsequent case counts in another age group $Y$, providing insights into directional transmission patterns. If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring]. As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger]. Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].

@@

\begin{align}
  \tag{2}
  T_{ag\,x \rightarrow ag\,y}(k,l) = \sum_{ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}} 
  p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) 
  \log \left(\frac{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}\right)}\right)
\end{align}

@@

Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:te)). For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (`r gam_results["0to4"]`). A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (`r gam_results["5to9"]`), where a decrease is observed during the first two weeks, followed by an increase. while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (`r gam_results["10to14"]`). For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`). Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.

@@

Knowledge about the transmission of the virus significantly improved over time as more studies have been published. While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate. It is now suggested that the impact of school openings on infection rates was limited, with children and teachers more often contracting infections through household or community contacts rather than within schools [@soriano2023policies]. This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups. While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups. Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization. In contrast, age groups ranging from 5 to 14 exhibit a stable profile after school closure. Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days. Thus, the results do not support the hypothesis that school closures were effective for group ages older than 5 years old. One possibility is that school closures may lead to increased informal social gatherings among children and teenagers outside of controlled school environments, potentially facilitating virus transmission. It is also possible that transmission within households intensified as children spent more time at home. These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.

The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups ($X$) and influenced age groups ($Y$). Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence. This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional. These findings highlight the importance of considering intergenerational interactions in designing effective control measures.

The entropy analysis revealed intriguing directional patterns of case counts between age groups. In some countries, cases among younger individuals appear predictive of cases among older individuals, while the opposite pattern is observed elsewhere. These differences may reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures. The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role. Future work should investigate these underlying mechanisms to better understand age-based transmission dynamics.

Furthermore, the quadrant analysis reveals similar patterns across multiple countries. Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts. On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts. However, these differences may also reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures. The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role.

The interpretation of our findings requires careful consideration of possible confounding factors. Changes in testing strategies, adherence to public health measures, and community transmission dynamics during school closures may have influenced the observed trends in COVID-19 cases. Our sensitivity analyses indicated that the overall trends remained robust after accounting for variations in testing rates and the presence of other public health interventions. However, these confounders may partially explain the heterogeneous effects observed across different age groups and countries. Future studies should explore these factors in greater detail to refine our understanding of the true impact of school closures.

In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures. While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.

@@
